Follow-up Samples

### 2016-12-06 submission/manuscript_outline_20161024.tex
temp changes

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

@@

\linenumbers

@@

Colorectal cancer (CRC) continues to be a worldwide health problem with
early detection being used as a key component in mitigating deaths due
to the disease. Previous research suggests that the bacterial microbiome
can be used as a biomarker for colorectal cancer (CRC). These reports
have mostly focused on investigating how the bacterial microbiome is
used at a single point in time to predict disease. In this study, we
assessed whether a model built with bacterial microbiome data could
accurately predict adenoma or carcinoma and adjust positive probability
for lesion (adenoma or carcinoma) after surgical treatment. This model
was tested on a \texttt{66} person group that included samples before
and after treatment to allow for the assessment of how the model adjusts
risk after treatment. The model chosen had a 10-fold cross validated AUC
of \texttt{0.819} and a test set AUC of \texttt{0.815}. : For the follow
up samples our Random Forest model significantly decreased the positive
probability of lesion compared to the initial samples for both adenoma
(P-value = \texttt{8.48e-07}) and carcinoma (P-value =
\texttt{5.96e-08}). In the test set, the prediction of lesion for the
initial samples had a sensitivity of \texttt{98.5}\% (\texttt{65/66})
while follow up it was \texttt{100}\% (\texttt{1\ /\ 1}). Our model
predicted that \texttt{39.4}\% of the 67-person cohort had normal colons
and \texttt{60.6}\% had a lesion. Our model suggests that treatment does
significantly reduce the probability of having a colonic lesion. Further
surveillance of these individuals will enable us to determine whether
models such as the one we present here can also be used to predict
recurrence of colorectal cancer.

@@

\textbf{\emph{Outcome of Model Training}} The range of the AUC for model
training ranged from a minimum of \texttt{0.795} to a maximum of
\texttt{0.854}. To be conservative the model chosen for prediction of
disease in the follow up samples had an AUC in the middle of all the
\texttt{100} runs which was \texttt{0.819}. Interestingly, the worst AUC
model from training performed the best on it's respective 80/20 split
test data {[}Figure 3{]}. In fact the 80/20 test performance showed that
the AUC for the middle model chosen was the most stable (best training
model test set AUC = \texttt{0.682}, middle training model test set AUC
= \texttt{0.815}, worse training model test set AUC = \texttt{0.931}).
That is to say it had the smallest change in AUC in comparison to the
other minimum and maximum AUC trained models. There was no significant
difference between the AUC of the best and middle training models
(P-value = \texttt{0.0698}). There was a significant difference between
the worse training model and both the best training model (P-value =
\texttt{0.000397}) and middle training model (P-value =
\texttt{0.0444}).

\textbf{\emph{Most Important Variables to the Model}} Overall, there
were a total of \texttt{32} variables identified as being present in
more than 50\% of the training models {[}Table S2{]}. The top 5 most
important bacterial OTUs were \texttt{Lachnospiraceae\ (Otu000013)},
\texttt{Escherichia/Shigella\ (Otu000018)},
\texttt{Ruminococcaceae\ (Otu000020)},
\texttt{Ruminococcus\ (Otu000017)}, and
\texttt{Porphyromonas\ (Otu000153)}. These 5 OTUs were present in at
least \texttt{99} out of the total \texttt{100} different 80/20 runs.

@@

There was \texttt{1} individual who still clearly had CRC on follow up
as well as \texttt{5} individuals whose status on follow up was unknown.
Although the \texttt{1} individual had a decrease in positive
probability their follow up sample was still higher than the cutoff
threshold of 0.5 (positive probability = \texttt{0.902}). Interestingly,
\texttt{2} individuals who were unknown on follow up still were over the
threshold cutoff of 0.5 even though, like the \texttt{1} individual with
clear CRC on follow up, the probability of an adenoma or carcinoma
decreased {[}Table S3{]}.

The follow up positive probabilities were not affected by either
chemotherapy treatment (uncorrected P-value = \texttt{0.583}) or
radiation therapy (uncorrected P-value = \texttt{0.809}). There was also
no difference in the amount of change in the positive probability based
on whether individuals received chemotherapy (unccorected P-value =
\texttt{0.224}) or radiation therapy (uncorrected P-value =
\texttt{0.846}).

@@

Naturally, it is curious that normal staples of many screening studies
such as Fusobacterium, Parvimonas, and Peptostreptococcus were not
present in the majority of the training models. One potential
explanation for this is that FIT provides the same information to the
model as these three organisms and so the model uses FIT preferentially
over them. This has been suggested to be the case in a previous study
(\textbf{insert Baxter Study}). Regardless, our study does not argue
against the importance of these bacterium in CRC initiation or
pathogenesis but rather that the best model does not utilize these
specific bacteria for prediction purposes.

@@

\textbf{\emph{Model Creation}} The Random Forest (\textbf{insert
citation}) algorithm was used to create the model used for prediction of
lesion (adenoma or carcinoma) for the \texttt{67} individuals with
follow up samples. The model included data on FIT, the bacterial
microbiome, sex, age, Body Mass Index (BMI), whether the indivdiual was
caucasion or not, history of cancer, and family history of cancer.
Non-binary data was checked for near zero variance and auto correlation.
Data columns that had near zero variance were removed. Columns that were
correlated with each other over a Spearman correlation coefficient of
0.75 had one of the two columns removed. This pre-processing was
performed with the R package caret (v\texttt{6.0.73}). The method of
choosing the best model involved taking the \texttt{490} samples and
making 100 80/20 (train/test) splits in the data where control and
lesion were equally represented in the 80 and 20 split, respectively.
This 80\% portion was then split again into an 80/20 split, and run
twenty times through 10-fold cross validation to optimzie the model's
Receiver Operator Characteristic (ROC) and tuneing both mtry and ntree.
This resulting model was then tested on the 20\% of the data that was
originally held out from this overall process. This was repeated 100
times with the best model after all the repeats chosen to be used on the
\texttt{67} samples with follow up. The default cutoff of a probability
of 0.5 was used as the threshold to classify individuals as having or
not having a lesion.

@@

\textbf{\emph{Statistical Analysis}} The R software package
(v\texttt{3.3.0}) was used for all statisitical analysis. Comparisons
between bacterial community structure utilized PERMANOVA (\textbf{insert
citation}) in the vegan package (v\texttt{2.4.1}) while comparisons
between ROC curves utilized the method by DeLong et al. (\textbf{insert
citation}) executed by the pROC (v\texttt{1.8}) package. Comparisons
between probabilities as well as overall amount of OTU between initial
and follow up samples utilized a paired wilcoxson ranked sum test. Where
multiple comparison testing was needed a Benjamini-Hochberg (BH)
correction was applied (\textbf{insert citation}) and a corrected
P-value of less than 0.05 was considered significant. Unless otherwise
stated the P-values reported are those of the BH corrected ones.

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\maketitle

@@

\newpage
\linenumbers

@@

Colorectal cancer (CRC) continues to be a worldwide health problem with
early detection being used as a key component in mitigating deaths due
to the disease. Previous research suggests that the bacterial microbiome
can be used as a biomarker for colorectal cancer (CRC). These reports
have mostly focused on investigating how the bacterial microbiome is
used at a single point in time to predict disease. In this study, we
assessed whether a model built with bacterial microbiome data could
accurately predict adenoma or carcinoma and adjust positive probability
for lesion (adenoma or carcinoma) after surgical treatment. This model
was tested on a \texttt{66} person group that included samples before
and after treatment to allow for the assessment of how the model adjusts
risk after treatment. The model chosen had a 10-fold cross validated AUC
of \texttt{0.819} and a test set AUC of \texttt{0.815}. : For the follow
up samples our Random Forest model significantly decreased the positive
probability of lesion compared to the initial samples for both adenoma
(P-value = \texttt{8.48e-07}) and carcinoma (P-value =
\texttt{5.96e-08}). In the test set, the prediction of lesion for the
initial samples had a sensitivity of \texttt{98.5}\% (\texttt{65/66})
while follow up it was \texttt{100}\% (\texttt{1 / 1}). Our model
predicted that \texttt{39.4}\% of the 67-person cohort had normal colons
and \texttt{60.6}\% had a lesion. Our model suggests that treatment does
significantly reduce the probability of having a colonic lesion. Further
surveillance of these individuals will enable us to determine whether
models such as the one we present here can also be used to predict
recurrence of colorectal cancer.

@@

\textbf{\emph{Outcome of Model Training}} The range of the AUC for model
training ranged from a minimum of \texttt{0.795} to a maximum of
\texttt{0.854}. To be conservative the model chosen for prediction of
disease in the follow up samples had an AUC in the middle of all the
\texttt{100} runs which was \texttt{0.819}. Interestingly, the worst AUC
model from training performed the best on it's respective 80/20 split
test data {[}Figure 3{]}. In fact the 80/20 test performance showed that
the AUC for the middle model chosen was the most stable (best training
model test set AUC = \texttt{0.682}, middle training model test set AUC
= \texttt{0.815}, worse training model test set AUC = \texttt{0.931}).
That is to say it had the smallest change in AUC in comparison to the
other minimum and maximum AUC trained models. There was no significant
difference between the AUC of the best and middle training models
(P-value = \texttt{NULL}). There was a significant difference between
the worse training model and both the best training model (P-value =
\texttt{NULL}) and middle training model (P-value = \texttt{NULL}).

\textbf{\emph{Most Important Variables to the Model}} Overall, there
were a total of \texttt{32} variables identified as being present in
more than 50\% of the training models {[}Table S2{]}. The top 5 most
important bacterial OTUs were \texttt{Bacteria (Otu000013)},
\texttt{Escherichia/Shigella (Otu000018)},
\texttt{Bacteria (Otu000020)}, \texttt{Ruminococcus (Otu000017)}, and
\texttt{Porphyromonas (Otu000153)}. These 5 OTUs were present in at
least \texttt{99} out of the total \texttt{100} different 80/20 runs.

@@

There was \texttt{1} individual who still clearly had CRC on follow up
as well as \texttt{5} individuals whose status on follow up was unknown.
Although the \texttt{1} individual had a decrease in positive
probability their follow up sample was still higher than the cutoff
threshold of 0.5 (positive probability = \texttt{0.927}). Interestingly,
\texttt{2} individuals who were unknown on follow up still were over the
threshold cutoff of 0.5 even though, like the \texttt{1} individual with
clear CRC on follow up, the probability of an adenoma or carcinoma
decreased {[}Table S3{]}.

The follow up positive probabilities were not affected by either
chemotherapy treatment (uncorrected P-value = \texttt{0.919}) or
radiation therapy (uncorrected P-value = \texttt{1}). There was also no
difference in the amount of change in the positive probability based on
whether individuals received chemotherapy (unccorected P-value =
\texttt{0.578}) or radiation therapy (uncorrected P-value =
\texttt{0.904}).

@@

Naturally, it is curious that normal staples of many screening studies
such as Fusobacterium, Parvimonas, and Peptostreptococcus were not
present in the majority of the training models. One potential
explanation for this is that FIT provides the same information to the
model as these three organisms and so the model uses FIT preferentially
over them. This has been suggested to be the case in a previous study
(\textbf{insert Baxter Study}). Regardless, our study does not argue
against the importance of these bacterium in CRC initiation or
pathogenesis but rather that the best model does not utilize these
specific bacteria for prediction purposes. Another potential reason why
we did not identify the usual suspects is that they may not change much
between our initial and follow up samples. That is to say that they are
consistently present even after removal of the lesion but surgery.

@@

\textbf{\emph{Model Creation}} The Random Forest (\textbf{insert
citation}) algorithm was used to create the model used for prediction of
lesion (adenoma or carcinoma) for the \texttt{67} individuals with
follow up samples. The model included data on FIT, the bacterial
microbiome, sex, age, Body Mass Index (BMI), whether the indivdiual was
caucasion or not, history of cancer, and family history of cancer.
Non-binary data was checked for near zero variance and auto correlation.
Data columns that had near zero variance were removed. Columns that were
correlated with each other over a Spearman correlation coefficient of
0.75 had one of the two columns removed. This pre-processing was
performed with the R package caret (v\texttt{6.0.72}). The method of
choosing the best model involved taking the \texttt{490} samples and
making 100 80/20 (train/test) splits in the data where control and
lesion were equally represented in the 80 and 20 split, respectively.
This 80\% portion was then split again into an 80/20 split, and run
twenty times through 10-fold cross validation to optimzie the model's
Receiver Operator Characteristic (ROC) and tuneing both mtry and ntree.
This resulting model was then tested on the 20\% of the data that was
originally held out from this overall process. This was repeated 100
times with the best model after all the repeats chosen to be used on the
\texttt{67} samples with follow up. The default cutoff of a probability
of 0.5 was used as the threshold to classify individuals as having or
not having a lesion.

@@

\textbf{\emph{Statistical Analysis}} The R software package
(v\texttt{3.3.2}) was used for all statisitical analysis. Comparisons
between bacterial community structure utilized PERMANOVA (\textbf{insert
citation}) in the vegan package (v\texttt{2.4.1}) while comparisons
between ROC curves utilized the method by DeLong et al. (\textbf{insert
citation}) executed by the pROC (v\texttt{1.8}) package. Comparisons
between probabilities as well as overall amount of OTU between initial
and follow up samples utilized a paired wilcoxson ranked sum test. Where
multiple comparison testing was needed a Benjamini-Hochberg (BH)
correction was applied (\textbf{insert citation}) and a corrected
P-value of less than 0.05 was considered significant. Unless otherwise
stated the P-values reported are those of the BH corrected ones.
